Conditions for successful wetting during collision between emulsion droplets and curved substrates

### 2012-10-24 Text/paper.tex
updated paper

\title{Conditions for successful wetting during collision between emulsion droplets and curved substrate}
\begin{document}
\maketitle
\section{Abstract}
\section{introduction}
Emulsion droplets are often used a delivery vehicle to deposit active ingredients on target substrates in application such as drug delivery, cosmetics and detergent products. In order to fulfill their function emulsion formulation are optimized in terms of emulsion droplets size, stability, surface properties and interaction forces with the target substrate. Virtually the properties of the emulsion droplets while stored before their use and their behavior after wetting the target substrate can be finely designed using well established physical models that identify the critical parameters controlling stability and adhesion properties.

However, to fulfill its function, the emulsion droplets need to be successfully transferred on the substrate. This transfer process, that involves the complex sequence of events that take the emulsion droplets from the formulated state to the substrate is not as well characterized and the role of the different parameters related to the emulsion, substrate and hydrodynamic interaction mediated by the continuous phase is still poorly defined. Recent AFM studied of emulsion interactions identified the critical role of hydrodynamic effect on the repulsion between two emulsions colliding at different speed. The effective disjointing pressure preventing contact and coalescence between the two emulsion droplets has an electrostatic and an hydrodynamic component. This suggest that also during the process of emulsion collision on a substrate successful contact and wetting of the surface critically depend on the interplay of the hydrodynamic and inter-facial properties.

@@

In the Next section we describe the idealized collision system simulated, the physical properties controlled in the simulation and the range of parameters explored (parameter space). In the following section we provide details about the simulation methods and its validation against other methodology previously proposed (Binary-liquid lattice Boltzmann model). Next we summarize the results of the parametric study focusing on the behavior of the emulsion as it collides with the substrate as a function of the interfacial and hydrodinamic parameters explored (Results). We then propose a map of the deposition efficiency within the explored parametric space (deposition efficiency map). Finally we summarize the results and relate to the related application areas (conclusions).

\section{parameter space}

\section{Binary-liquid lattice Boltzmann model}
The lattice Boltzmann equation (LBE) operates on a rectangular grid representing the
physical domain. It utilizes
probability distribution functions (also known as particle populations)
containing information about
macroscopic variables, such as fluid density, momentum, and the phase order parameter for multiphase models. LBE consists of
two parts: a local collision step, and a propagation step which transports
information from one node to another along some 
directions specified by the discrete velocity set.
The LBE is typically implemented as follows:
\begin{equation}
\label{standard:implementation}
\begin{aligned}
&f_i^{*}(\bm{x},t)=\omega f_i^{eq}(\bm{x},t)-(1-\omega) f_i(\bm{x},t) +
F_i,&&\text{ collision step}\\
&f_i(\bm{x}+\bm{c_i},t+1)=f_i^{*}(\bm{x},t),&&\text{ propagation step}, 
\end{aligned}
\end{equation}
where $f_i$ is the probability distribution function in the direction $\bm{c_i}$,
 $f_i^{eq}$ is the equilibrium probability distribution function, $\omega$ is the
relaxation parameter, and $F_i$ is the external force population. The force population
represents an external physical force and is implemented in the current work using the scheme
outlined in \citet{guo}.

The binary fluid LB model is
based on a free-energy functional \cite{swift,landau}, and operates with two
sets of populations: one to track the pressure and the velocity fields, and another to represent the
phase field $\phi$ indicating the gas or liquid.

The model we use is a two-dimensional nine-velocity (D2Q9) model,
with equilibrium populations \cite{pooley-contact}:
\begin{equation}
\label{set:equilibrium:binary}
\begin{aligned}
&f_i^{eq}&&=w_i 
\biggl(3
p_0 - k \phi \Delta \phi
+\rho\frac{u_{\alpha}c_{i\alpha}}{c_s^2}+\rho \frac{Q_{i\alpha\beta}u_{\alpha } u_ {
\beta}}{2 c_s^4}\biggr)\\
&&&+k\bigl(w_i^{xx} (\partial_x \phi)^2+w_i^{yy} (\partial_y \phi)^2 +w_i^{xy} \partial_x
\phi \partial_y \phi \bigr), 1\leq i \leq 8\\
&f_0^{eq}&&=\rho-\sum_{i\neq0}{f_i^{eq}}\\
&g_i^{eq}&&=w_i\left(\Gamma \mu + \phi\frac{ c_{i\alpha} u_{i\alpha}}{c_s^2}+\phi
\frac{Q_{i\alpha\beta}u_{\alpha}u_{\beta}}{2 c_s^4}\right), 1\leq i \leq 8 \\
&g_0^{eq}&&=\phi-\sum_{i\neq0}{g_i^{eq}}\quad,
\end{aligned}
\end{equation}
where $\Gamma$ is the mobility parameter; the chemical potential
$\mu=-A\phi+A\phi^3-k\Delta\phi$; $k$ is the parameter related to the surface
tension; $A$ is the parameter of the free-energy model. The bulk pressure
is expressed as $p_0=c_s^2 \rho +A (-0.5 \phi^2+0.75 \phi^4)$ with
the sound speed $c_s^2=1/3$. 
Parameters specific to the D2Q9 grid are the weights
$w_i=\left\{\frac{4}{9},\frac{1}{9},\frac{1}{9},\frac{1}{9},\frac{1}{9},
\frac{1}{36},\frac{1}{36},\frac{1}{36},\frac{1}{36}\right\}$, and the tensor
$Q_{i\alpha\beta}=c_{i\alpha} c_{i\beta} - c_s^2 \delta_{\alpha\beta}$.  
Other weights 
are as follows:
$w^{xx}_{1-2}=w^{yy}_{3-4}=1/3$, $w^{xx}_{3-4}=w^{yy}_{1-2}=-1/6$,
$w^{xx}_{5-8}=w^{yy}_{5-8}=-1/24$, $w^{xy}_{1-4}=0$, $w^{xy}_{5-6}=1/4$ and
$w^{xy}_{7-8}=-1/4$.

@@

\begin{figure}
\includegraphics[width=\textwidth]{Figures/free_surface.eps}
\caption{Free-surface boundary condition represented in the lattice Boltzmann method. 
Boundary nodes are depicted by crosses, and fluid nodes are represented by dots. One imposes only phase values at the nodes to calculate laplacians used in the equilibrium functions. In the population space, one uses the bounce-back rule \cite{yu}, which conserves mass.
\label{fig:free:surface}}
\end{figure}

\begin{figure}
\includegraphics[width=0.5\textwidth]{Figures/droplet_shapes_curved_substrates.eps}
\includegraphics[width=0.5\textwidth]{Figures/main_curve_circle.eps}\\
\caption{Equilibrium angles at the curved substrate. One can see that the contact angle at the curved substrate mimics the analytically calculated contact angle at the flat substrate. \label{fig:equilibrium:droplet:curved}}
\end{figure}

\section{Results}

\section{Deposition efficiency map}

\section{Conclusions}
The work presented investigates the interaction between emulsion droplets and curved substrates to understand the critical factors controlling the efficiency and the kinetics of the wetting process during collision. We use 2D Lattice Boltzmann methods to simulate spherical emulsion droplets colliding on a curved "cylindrical" substrate. The properties of the substrate are set by controlling its curvature and the three phase contact angle with the emulsion droplet and the surrounding fluid. The emulsion droplet is characterized by its size, viscosity and interfacial tension with the surrounding fluid. As expected we observe that the curvature of the droplet or substrate and the kinetics of the collision process does not effect the equilibrium three phase contact angle after wetting. However a detailed analysis of the collision process reveals that droplets deformation at wetting critically depends on the collision speed and relative dimension of droplet and substrate. Large droplets approaching the substrate very slowly deform very mildly before wetting while during fast collision a higher deformation is achieved before wetting occurs. For small droplets the same trend is found but overall smaller deformation are observed. The results can be rationalized if we consider a hydrodynamic disjoining pressure that retard contact between droplets and substrate. Based on this finding we can map the optimal conditions to achieve successful wetting, deposition and retention of an emulsion droplet on a substrate. This results can be used to guide the design of emulsions used as actives delivery vehicles. Emulsion size and interfacial properties can be tuned to optimize deposition efficiency on the relevant substrates based on the hydrodynamic parameters of the collision process. Such information are of important practical utility in pharmaceutical and cosmetic application in which emulsion droplets are often used to deposit active ingredient on tissues or fibers.  
\bibliographystyle{unsrt}
\bibliography{paper}
\end{document}

>>>>

\title{Conditions for successful wetting during collision between emulsion droplets and curved substrates}
\begin{document}
\maketitle
\section{Abstract}
In consumer products, cosmetics, and medicines many benefits are delivered by deposition of emulsion droplets onto hair, skin, or other tissues.
The interaction of emulsion droplets with solid substrates is critical in determining the efficiency of delivery and can often be studied using the experiments with a model substrate like a fiber mesh. While numerous properties of emulsion droplets, like stability, size, interfacial tension, and adhesion can be varied, little is known about how each parameter influences the collision, wetting, and retention probability.  The complex hydrodynamic interactions at play during the droplet deposition process must also be considered in addition to materials properties. In this work we use Lattice Boltzmann, LB, simulation methods to investigate such problems. The simultaneous need to simulate a multiphase system and flow through complex geometries makes LB simulation a good candidate for a parametric exploration of the effects of formulation and process parameters on deposition efficiency. We find that the condition for succesfull wetting upon collision critically depend on material properties such as relative size of droplet and substrate, interfacial tension and viscosity ratio. Additionally we find that also the hydrodinamic conditions play an important role to that can be rationalized considering the presence of an additional hydrodynamic disjoining pressure that depends on the collision speed. Based on this finding we can map the optimal conditions to achieve successful wetting, deposition and retention of an emulsion droplet on a substrate. This results can be used to guide the design of emulsions used as actives delivery vehicles.

\pagebreak
\section{introduction}
Emulsion droplets are often used as delivery vehicle to deposit active ingredients on target substrates in application such as drug delivery, cosmetics and detergent products. To fulfill their function, emulsion formulation are optimized in terms of emulsion droplets size, stability, surface properties and interaction forces with the target substrate. Virtually the properties of the emulsion droplets while stored before their use and their behavior after wetting the target substrate can be finely designed using well established physical models that identify the critical parameters controlling stability and adhesion properties. However, the emulsion droplets need to be successfully transferred on the substrate and little is known on how to optimize this part of the delivery process. The deposition of the emulsion droplet involves the complex sequence of events that take the emulsion droplets from the formulated state to the substrate.  This transfer process is not well characterized and the role of the different parameters related to the emulsion, substrate and hydrodynamic interaction mediated by the continuous phase is still poorly understood. Recent AFM studied of emulsion interactions identified the critical role of hydrodynamic effect on the repulsion between two emulsions colliding at different speed. The effective disjointing pressure preventing contact and coalescence between the two emulsion droplets has an electrostatic and an hydrodynamic component. This suggest that also during the process of emulsion collision on a substrate successful contact and wetting of the surface critically depend on the interplay of the hydrodynamic and inter-facial properties.

@@

In the Next section we describe the idealized collision system simulated, the physical properties controlled in the simulation and the range of parameters explored (parameter space). In the following section we provide details about the simulation methods and its validation against other methodology previously proposed (Binary-liquid lattice Boltzmann model). Next we summarize the results of the parametric study focusing on the behavior of the emulsion as it collides with the substrate as a function of the interfacial and hydrodinamic parameters explored (Results). We then propose a map of the deposition efficiency within the explored parametric space (deposition efficiency map). Finally we summarize the results and discuss their relevance in the related application areas (conclusions).

\pagebreak 
\section{parameter space}
While the detail of the simulation method will be given in the next section, here we describe the simple physical system that we simulate and the parameters explored.

A droplet, dispersed phase, of dimension $R_e$ is placed at the beginning of a channel of fixed width $l_c$ and length $L_c$. The droplet is surrounded by a continuous liquid phase that fills the entire channel. In the simulation we can vary the relative viscosity between dispersed and continuous phase $\eta_r$ and the interfacial tension between the two phases $\gamma$.

In the middle of the channel, downstream from the droplet position, we place a solid curved substrate with a circular cross section of radius $R_s$. The properties of the substrate surface can be tune to obtain a range of equilibrium three phase contact angles between the dispersed phase, continuous phase, and substrate surface $\eta_s$.

At the beginning of the simulation, during the equilibration step,  a body force field is applied to the continuous phase to generate flow in the channel, while the flow field develops toward a steady state the droplet is held in the original position. The amplitude of the body force field can be varied to generate a range of hydrodynamic conditions for the collision process.

Once the continuous phase flow is established the dispersed phase droplet is released and accelerates under the action of the body force and viscous forces toward the substrate, acceleration step. Since the droplet is placed in the middle of the channel relative to its width as for the substrate, the induced collision is a central collision.

We describe the collision as a two step process. Initially during the collision pre-wetting stage, the droplets fell the presence of the substrate, applies a pressure on it, and deforms as the liquid layer of continuous phase that separate the droplet and substrate surface get depleted.

Eventually contact between the two surfaces is produced, collision wetting stage, and the droplet wet the surface to reach an equilibrium position that depend on the balance between the interfacial forces and the hydrodynamic interaction with the continuous phase still in motion.

Snapshots form a sumulation result representing the different step are represented in figure \ref{fig:simulation:explanation}

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{Figures/collision_simulation.eps}
\caption{Steps of the simulation of the collision process\label{fig:simulation:explanation}}
\end{figure}

\pagebreak 
\section{Binary-liquid lattice Boltzmann model}
The lattice Boltzmann equation (LBE) operates on a rectangular grid representing the physical domain. It utilizes probability distribution functions (also known as particle populations) containing information about macroscopic variables, such as fluid density, momentum, and the phase order parameter for multiphase models. LBE consists of two parts: a local collision step, and a propagation step which transports information from one node to another along some 
directions specified by the discrete velocity set. The LBE is typically implemented as follows:
\begin{equation}
\label{standard:implementation}
\begin{aligned}
&f_i^{*}(\bm{x},t)=\omega f_i^{eq}(\bm{x},t)-(1-\omega) f_i(\bm{x},t) +
F_i,&&\text{ collision step}\\
&f_i(\bm{x}+\bm{c_i},t+1)=f_i^{*}(\bm{x},t),&&\text{ propagation step}, 
\end{aligned}
\end{equation}
where $f_i$ is the probability distribution function in the direction $\bm{c_i}$,
 $f_i^{eq}$ is the equilibrium probability distribution function, $\omega$ is the relaxation parameter, and $F_i$ is the external force population. The force population represents an external physical force and is implemented in the current work using the scheme outlined in \citet{guo}.

The binary fluid LB model is based on a free-energy functional \cite{swift,landau}, and operates with two sets of populations: one to track the pressure and the velocity fields, and another to represent the phase field $\phi$ indicating the gas or liquid. The model we use is a two-dimensional nine-velocity (D2Q9) model, with equilibrium populations \cite{pooley-contact}:
\begin{equation}
\label{set:equilibrium:binary}
\begin{aligned}
&f_i^{eq}&&=w_i 
\biggl(3
p_0 - k \phi \Delta \phi
+\rho\frac{u_{\alpha}c_{i\alpha}}{c_s^2}+\rho \frac{Q_{i\alpha\beta}u_{\alpha } u_ {
\beta}}{2 c_s^4}\biggr)\\
&&&+k\bigl(w_i^{xx} (\partial_x \phi)^2+w_i^{yy} (\partial_y \phi)^2 +w_i^{xy} \partial_x
\phi \partial_y \phi \bigr), 1\leq i \leq 8\\
&f_0^{eq}&&=\rho-\sum_{i\neq0}{f_i^{eq}}\\
&g_i^{eq}&&=w_i\left(\Gamma \mu + \phi\frac{ c_{i\alpha} u_{i\alpha}}{c_s^2}+\phi
\frac{Q_{i\alpha\beta}u_{\alpha}u_{\beta}}{2 c_s^4}\right), 1\leq i \leq 8 \\
&g_0^{eq}&&=\phi-\sum_{i\neq0}{g_i^{eq}}\quad,
\end{aligned}
\end{equation}
where $\Gamma$ is the mobility parameter; the chemical potential
$\mu=-A\phi+A\phi^3-k\Delta\phi$; $k$ is the parameter related to the surface tension; $A$ is the parameter of the free-energy model. The bulk pressure is expressed as $p_0=c_s^2 \rho +A (-0.5 \phi^2+0.75 \phi^4)$ with the sound speed $c_s^2=1/3$. 
Parameters specific to the D2Q9 grid are the weights $w_i=\left\{\frac{4}{9},\frac{1}{9},\frac{1}{9},\frac{1}{9},\frac{1}{9},
\frac{1}{36},\frac{1}{36},\frac{1}{36},\frac{1}{36}\right\}$, and the tensor
$Q_{i\alpha\beta}=c_{i\alpha} c_{i\beta} - c_s^2 \delta_{\alpha\beta}$.  
Other weights are as follows:
$w^{xx}_{1-2}=w^{yy}_{3-4}=1/3$, $w^{xx}_{3-4}=w^{yy}_{1-2}=-1/6$,
$w^{xx}_{5-8}=w^{yy}_{5-8}=-1/24$, $w^{xy}_{1-4}=0$, $w^{xy}_{5-6}=1/4$ and
$w^{xy}_{7-8}=-1/4$.

@@

\begin{figure}[h!]
\includegraphics[width=0.5\textwidth]{Figures/free_surface.eps}
\caption{Free-surface boundary condition represented in the lattice Boltzmann method. 
Boundary nodes are depicted by crosses, and fluid nodes are represented by dots. One imposes only phase values at the nodes to calculate laplacians used in the equilibrium functions. In the population space, one uses the bounce-back rule \cite{yu}, which conserves mass.
\label{fig:free:surface}}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=0.5\textwidth]{Figures/droplet_shapes_curved_substrates.eps}
\includegraphics[width=0.5\textwidth]{Figures/main_curve_circle.eps}\\
\caption{Equilibrium angles at the curved substrate. One can see that the contact angle at the curved substrate mimics the analytically calculated contact angle at the flat substrate. \label{fig:equilibrium:droplet:curved}}
\end{figure}

\pagebreak
\section{Results}

Figure reference \ref{fig:droplet:shapes:curved:substrate:F1} \ref{fig:droplet:shapes:curved:substrate:F5} \ref{fig:extra:pressure}

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{Figures/shapes_F1_Grad-20.eps}
\caption{Droplet shape at simulation time step before succesful wetting for flow conditions characterized by body force paramerer F1 and contact angle conditions characterized by gradient parameter -20. A range of different initial droplet radius is shown (R=10,20,30,40,50) while the subtrate radius was kept constant (R=20)  \label{fig:droplet:shapes:curved:substrate:F1}}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{Figures/shapes_F5_Grad-20.eps}
\caption{Droplet shape at simulation time step before succesful wetting for flow conditions characterized by body force paramerer F5 and contact angle conditions characterized by gradient parameter -20. A range of different initial droplet radius is shown (R=10,20,30,40,50) while the subtrate radius was kept constant (R=20)  \label{fig:droplet:shapes:curved:substrate:F5}}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{Figures/comparison_another_approach.eps}
\caption{Extra pressure inside emulsion droplets measured at the timestep just before weeting the substrate for flow conditions characterized by body force parameter F1 full circles and F5 triagles. The solid line indicates the theoretical value for the laplace pressure for the different emulsion sizes.\label{fig:extra:pressure}}
\end{figure}

\pagebreak
\section{Deposition efficiency map}

Figure reference \ref{fig:deposition:map}

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{Figures/results_map.eps}
\caption{Deposition map.  Label 1 stays for 1 droplet attached front, 2 stays for 1 droplet attached back, 3 stays for 2 droplets attached back, 4 stays for not attached, 5 stays for breakup to 2 droplets, 6 stays for breakup with coalescense\label{fig:deposition:map}}
\end{figure}

\pagebreak
\section{Conclusions}
The work presented investigates the interaction between emulsion droplets and curved substrates to understand the critical factors controlling the efficiency and the kinetics of the wetting process during collision. We use 2D Lattice Boltzmann methods to simulate spherical emulsion droplets colliding on a curved "cylindrical" substrate. The properties of the substrate are set by controlling its curvature and the three phase contact angle with the emulsion droplet and the surrounding fluid. The emulsion droplet is characterized by its size, viscosity and interfacial tension with the surrounding fluid. As expected we observe that the curvature of the droplet or substrate and the kinetics of the collision process does not effect the equilibrium three phase contact angle after wetting. However a detailed analysis of the collision process reveals that droplets deformation at wetting critically depends on the collision speed and relative dimension of droplet and substrate. Large droplets approaching the substrate slowly deform very mildly before wetting while during fast collision a higher deformation is achieved before wetting occurs. For small droplets the same trend is found but overall smaller deformation are observed. The results can be rationalized if we consider a hydrodynamic disjoining pressure that retard contact between droplets and substrate. Based on this finding we can map the optimal conditions to achieve successful wetting, deposition and retention of an emulsion droplet on a substrate. This results can be used to guide the design of emulsions used as actives delivery vehicles. Emulsion size and interfacial properties can be tuned to optimize deposition efficiency on the relevant substrates based on the hydrodynamic parameters of the collision process. Such information are of important practical utility in pharmaceutical and cosmetic application in which emulsion droplets are often used to deposit active ingredient on tissues or fibers.  
\bibliographystyle{unsrt}
\bibliography{paper}
\end{document}
